---
title: "weeping mulla mulla: Ptilotus calostachyus"
author: "ACDC: Minh Pham, Kelly Nguyen"
format: html
---

```{r libraries, echo=False}
library(galah)
library(tidyverse)
library(kableExtra)
```


## Introduction

The *Ptilotus calostachyus F.Muell*, which is also known as the **Weeping Mulla Mulla**, is an endemic plant that is commonly found in the Western Australia territory(1). This species thrives in arid region, particular in sandy habitats with red and stony plains(1).  Being a *perennial* plant, it can live up to more than two years with the seasonal pattern of being dormant in the winter and sprouting again in springtime - (1). Moreover, **Weeping Mulla Mulla** displays erect or spreading, herb or shrub characteristics, meaning that it can be found as a tall plant and/or covering an area, as well as with/without a woody body. They can grow from 0.2 to 2 meters tall with pink-white flowers, blooming from March to October. Despite the widespread of its coverage, there are little to none organized sighting campaign of the **Weeping Mulla Mulla** due to the outback nature of its habitat and lesser popularity as compared to other iconic Australia ecotourism species. 

Five expectations when exploring the **Weeping Mulla Mulla** sighting:
- We expect most of the sightings to occur between March to October.
- We expect most of the sighting location to be in the Western Australia territory.
- We expect the sightings to be consistent numbers of sightings with an increasing trend as sighting/reporting/monitoring methods become better. 

## Data description

<!--
2. Download all the occurrence records from the Atlas of Living Australia, ideally scripted using the `atlas_occurrences()` function from the `galah` package. Make sure you include date, time, collector identification details. Save this into a `data-raw` folder in your repo.
-->
```{r , eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
galah_config(email = "mpha0054@student.monash.edu",
             download_reason_id = 10, 
             verbose = TRUE)

raw_data <- ala_occurrences(
  taxa = select_taxa("Ptilotus calostachyus"))

save(raw_data, file = "data-raw/raw-data.rda")
```


## Initial data analysis

<!--
3. Perform a data quality check, and then go ahead and clean the data, saving only the wild sightings. Save the resulting data in a `data` folder as an `WILD_THING_COMMON_NAME.rda` file, so that any special variable types are clearly available. (Replace `WILD_THING_COMMON_NAME` with lower case common name of your assigned wild thing.) *Note:* Some are museum or historical records or live in an animal sanctuary or zoo, which do not truly capture their wild behaviour. Downloading the event information (`galah_select(group = "event")`) can help identify wild sightings.

Weather data pre-processing here too.
-->

### Processing the data. 

* This is to differentiate between wild sightings and collected sample in museums.

```{r}
galah_config(email = "mpha0054@student.monash.edu",
             download_reason_id = 10, 
             verbose = TRUE)

storage <- galah_call() %>%
  galah_identify("Weeping mulla mulla") %>%
  galah_select(basisOfRecord , decimalLatitude, decimalLongitude,eventDate, scientificName, taxonConceptID, recordID, dataResourceName, eventRemarks, eventTime, eventID, eventDate, samplingEffort, samplingProtocol, group = c("basic", "event")) %>% 
  atlas_occurrences() %>%
  filter(basisOfRecord == "HUMAN_OBSERVATION")
```



## Exploratory data analysis

<!--
4. Make plots of the data to check each of your expectations, and report what you learn. *Note:* This might also be a place where visual inference methods *may* be useful to determine if what is seen agrees or disputes what you expected. 

Weather data exploration here too.
-->

## Summary 

## References
