---
title: "Weeping Mulla Mulla: Ptilotus Calostachyus"
author: "**Team name:** ACDC **Team members:** Khuyen Nguyen kngu0086@student.monash.edu & Khanh Pham"
format: html
---

## Introduction

 - We expect the species to thrive in dry conditions, with temperatures ranging from approximately 24 to 34 degrees Celsius _(TravelOnline Australia)_.
 - We anticipate that there may have been fewer sightings during the COVID-19 pandemic. Since the data relies on volunteer work, there might have been fewer sightings due to a lack of reporting during this time.

## Data description

```{r species-data, message=FALSE}
library(kableExtra)
library(galah)
library(spData)
library(tidyverse)
galah_config(email = "kngu0086@student.monash.edu",
             download_reason_id = 10,
             verbose = FALSE)
rawdata <- galah_call() %>% 
  galah_identify("Ptilotus calostachyus")  %>% 
  atlas_occurrences()
save(rawdata, file = "data-raw/rawdata.rda")
```

The data is sourced from the Atlas of Living Australia website, which aggregates multiple data sources and includes variables indicating location, date, time, scientific name, resource names, and species status. There are `r nrow(rawdata)` observations and `r ncol(rawdata)` variables in the raw data set.

For our analysis, we will primarily utilise the location and event time data. However, a comprehensive dictionary is provided below:

```{r data-dictionary, echo=FALSE, warning=FALSE}
data <- data.frame(
  Variable = c("decimalLatitude", "decimalLongitude", "eventDate", "scientificName", "taxonConceptID", "recordID", "dataResourceName",  "occurrenceStatus" ),
  Description = c(
    "Latitude value of sighting in decimal format.",
    "Longitude value of sighting in decimal format.",
    "Recorded sighting date and time in the format YYYY-MM-DD, hh:mm:ss. In some cases, no time was recorded, and therefore, it is left as 00:00:00.",
    "The scientific name of the species: Ptilotus Calostachyus.",
    "The taxonomy website ID: https://id.biodiversity.org.au/node/apni/2910906.",
    "The record ID of each observation.",
    "The data resource of the observation of the 23 [data partners](https://bie.ala.org.au/species/https://id.biodiversity.org.au/node/apni/2910906#data-partners).",
    "The occurrence status: Present."
  )
)

kable(data, format = "markdown")
```

## Initial data analysis

```{r species-data-cleaning}
weeping_mulla_mulla <- rawdata %>%
  select(decimalLongitude, decimalLatitude, eventDate) %>%
  rename(Longitude = decimalLongitude,
         Latitude = decimalLatitude) %>%
  mutate(
    eventDate = as.Date(eventDate),
    Year = lubridate::year(eventDate),
    Month = lubridate::month(eventDate),
    monthName = lubridate::month(eventDate, label = TRUE, abbr = TRUE),
    Day = lubridate::day(eventDate)
  ) %>%
  filter(!is.na(eventDate)) %>%
  filter(!is.na(Longitude))
save(weeping_mulla_mulla, file = ("data/weeping_mulla_mulla.rda"))
```

```{r weather-data, warning=FALSE, message=FALSE}
library(rnoaa)
aus_stations <- ghcnd_stations() |>
  filter(str_starts(id, "ASN")) |>
  filter(last_year >= 2020) |>
  mutate(wmo_id = as.numeric(wmo_id),
         name = str_to_lower(name)) |>
  select(-state, -gsn_flag) |>
  filter(element %in% c("PRCP", "TMAX", "TMIN")) |>
  nest(element: last_year) |>
  rowwise() |>
  filter(nrow(data) == 3) |>
  select(-data) 
```

```{r, warning=FALSE, message=FALSE}
narrogin <- aus_stations |>
  filter(id == "ASN00010614") |>
  rowwise() |>
  mutate(ts = list(meteo_pull_monitors(
    monitors = id, var = c("PRCP", "TMAX", "TMIN"),
    date_min = "2004-01-01",
    date_max = "2023-09-13") |>
      select(-id))) |>
  rename(lat = latitude, long = longitude, elev = elevation) |> select(id, long, lat, elev, name, wmo_id, ts) %>%
  unnest(ts) |>
  mutate(tmax = tmax/10, tmin = tmin/ 10)
```

## Exploratory data analysis

```{r}
data(world, package = "spData")
australia <- world[world$iso_a2 == "AU",]

australia_plot <- ggplot(australia) + geom_sf(fill = "#5b9b2e")

combined_plot <- australia_plot +
  geom_point(
    data = weeping_mulla_mulla,
    aes(x = Longitude, y = Latitude),
     colour = "#c599b2", 
) +
  labs(title = "Weeping Mulla Mulla Sightings Across Australia") +
  theme_minimal() 

combined_plot
```
```{r}
combined_plot +
  facet_wrap(~monthName, ncol = 6) +
  theme_void() +
  labs(title = "Weeping Mulla Mulla Sightings By Month") 
```





## Summary 

## References

 - TravelOnline Australia Pty Ltd. Western Australia Weather & Climate. Retrieved September 27, 2023, from https://www.travelonline.com/western-australia/weather
 - The Atlas of Living Australia. Ptilotus calostachyus F.Muell.Retrieved September 27, 2023, from https://bie.ala.org.au/species/https://id.biodiversity.org.au/node/apni/2910906#overview


