---
title: "weeping mulla mulla: Ptilotus calostachyus"
author: "ACDC: Minh Pham, Kelly Nguyen"
format: html
editor: 
  markdown: 
    wrap: 72
---

```{r libraries, echo=False}
library(galah)
library(tidyverse)
library(kableExtra)
```

## Introduction

The *Ptilotus calostachyus F.Muell*, which is also known as the
**Weeping Mulla Mulla**, is an endemic plant that is commonly found in
the Western Australia territory (Florabase n. d.). This species thrives
in arid region, particular in sandy habitats with red and stony
plains(1). Being a *perennial* plant, it can live up to more than two
years with the seasonal pattern of being dormant in the winter and
sprouting again in springtime - (1). Moreover, **Weeping Mulla Mulla**
displays erect or spreading, herb or shrub characteristics, meaning that
it can be found as a tall plant and/or covering an area, as well as
with/without a woody body. They can grow from 0.2 to 2 meters tall with
pink-white flowers, blooming from March to October. Despite the
widespread of its coverage, there are little to none organized sighting
campaign of the **Weeping Mulla Mulla** due to the outback nature of its
habitat and lesser popularity as compared to other iconic Australia
ecotourism species.

Five expectations when exploring the **Weeping Mulla Mulla** sighting: -
We expect most of the sightings to occur between March to October. - We
expect most of the sighting location to be in the Western Australia
territory. - We expect the sightings to be consistent numbers of
sightings with an increasing trend as sighting/reporting/monitoring
methods become better.

## Data description

```{=html}
<!--
2. Download all the occurrence records from the Atlas of Living Australia, ideally scripted using the `atlas_occurrences()` function from the `galah` package. Make sure you include date, time, collector identification details. Save this into a `data-raw` folder in your repo.
-->
```
```{r , eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
galah_config(email = "mpha0054@student.monash.edu",
             download_reason_id = 10, 
             verbose = TRUE)

raw_data <- ala_occurrences(
  taxa = select_taxa("Ptilotus calostachyus"))

save(raw_data, file = "data-raw/raw-data.rda")
```

## Initial data analysis

```{=html}
<!--
3. Perform a data quality check, and then go ahead and clean the data, saving only the wild sightings. Save the resulting data in a `data` folder as an `WILD_THING_COMMON_NAME.rda` file, so that any special variable types are clearly available. (Replace `WILD_THING_COMMON_NAME` with lower case common name of your assigned wild thing.) *Note:* Some are museum or historical records or live in an animal sanctuary or zoo, which do not truly capture their wild behaviour. Downloading the event information (`galah_select(group = "event")`) can help identify wild sightings.

Weather data pre-processing here too.
-->
```
### Processing the data.

This is to differentiate between wild sightings and collected sample in museums. We have also checked the `BASIS_OF_RECORD_INVALID` to determine the reliability of each observation, in which all the `TRUE` results were removed in the process of sub-setting `HUMAN_OBSERVATION` entry in ``basisOfRecord`.
  

```{r}
galah_config(email = "mpha0054@student.monash.edu",
             download_reason_id = 10, 
             verbose = TRUE)

storage <- galah_call() %>%
  galah_identify("Weeping mulla mulla") %>%
  galah_select(basisOfRecord , decimalLatitude, decimalLongitude,eventDate, scientificName, taxonConceptID, recordID, dataResourceName, eventRemarks, eventTime, eventID, eventDate, samplingEffort, samplingProtocol, group = c("basic", "event","assertions")) %>% 
  atlas_occurrences() %>%
  filter(basisOfRecord == "HUMAN_OBSERVATION")
```

```{r species-data-cleaning}
weeping_mulla_mulla <- storage %>%
  select(decimalLongitude, decimalLatitude, eventDate) %>%
  rename(longitude = decimalLongitude,
         latitude = decimalLatitude) %>%
  mutate(
    eventDate = as.Date(eventDate),
    Year = lubridate::year(eventDate),
    Month = lubridate::month(eventDate),
    monthName = lubridate::month(eventDate, label = TRUE, abbr = TRUE),
    Day = lubridate::day(eventDate)
  ) %>%
  filter(!is.na(eventDate)) %>%
  filter(!is.na(longitude))
```

### Checking data quality

We will be using `visdat` library to check for data type errors

```{r message=FALSE, warning=FALSE}
library(visdat)

vis_dat(weeping_mulla_mulla)
```

**Discussion**
The result of `visdat` does not signify any anomalies that needed to be addressed before proceeding with further analysis.

```{r message=FALSE, warning=FALSE}
vis_guess(weeping_mulla_mulla)
```

**Discussion**
`vis_guess` shows that the *longitude* and *latitude* columns both have a record with potentially integer coordinates entry. While this is a sign of rounding given that the coordinates span from a length of values with 5 decimal places precision, we believe this value does not need any alterations as sighting coordinates span across a very wide geographical area, making rounding to the nearest whole number (1.0 degree is 111 kilometers) insignificant in our spatial analysis.

### Weather data

```{r weather-data, warning=FALSE, message=FALSE}
library(rnoaa)
aus_stations <- ghcnd_stations() |>
  filter(str_starts(id, "ASN")) |>
  filter(last_year >= 2020) |> 
  mutate(wmo_id = as.numeric(wmo_id),
         name = str_to_lower(name)) |>
  select(-state, -gsn_flag) |>
  filter(element %in% c("PRCP", "TMAX", "TMIN")) |>
  nest(element: last_year) |>
  rowwise() |>
  filter(nrow(data) == 3) |>
  select(-data) 
```


The distribution of the Weeping Mulla Mulla commonly spans the local
government areas (LGAs) of Ashburton, Broome, Derby-West Kimberley, East
Pilbara, Halls Creek, Karratha, Laverton, Meekatharra, Ngaanyatjarraku,
Port Hedland, Upper Gascoyne, Wyndham-East Kimberley, as per
([Florabase](https://florabase.dbca.wa.gov.au/browse/profile.php/2704)).

As such, we can subset the temperature data for some of these specific
suburbs.

* broome airport ASN00003003 
* derby aero ASN00003032 
* halls creek airport ASN00002012 
* karratha aero ASN00004083 
* laverton aero ASN00012305 
* laverton raaf ASN00087031 
* meekatharra airport ASN00007045 
* port hedland airport ASN00004032 
* gascoyne junction ASN00006022 
* wynyard airport ASN00091107
* wyndham aero ASN00001006

```{r, warning=FALSE, message=FALSE}
broomeairport <- aus_stations |>
  filter(id == "ASN00003003") |>
  rowwise() |>
  mutate(ts = list(meteo_pull_monitors(
    monitors = id, var = c("PRCP", "TMAX", "TMIN"),
    date_min = "2004-01-01",
    date_max = "2023-09-13") |>
      select(-id))) |>
  rename(lat = latitude, long = longitude, elev = elevation) |> select(id, long, lat, elev, name, wmo_id, ts) %>%
  unnest(ts) |>
  mutate(tmax = tmax/10, tmin = tmin/ 10)
```

## Exploratory data analysis

**Spatial expectation**

We expect most of the sighting locations to be in Western Australia
territory. The plot of sightings below has met our expectations;
however, there are also many sightings in the Northern Territory.

```{r}
data(world, package = "spData")
australia <- world[world$iso_a2 == "AU", ]

australia_plot <- ggplot(australia) + geom_sf(fill = "#5b9b2e")

combined_plot <- australia_plot +
  geom_point(data = weeping_mulla_mulla,
             aes(x = longitude, y = latitude),
             colour = "#c599b2",) +
  labs(title = "Weeping Mulla Mulla Sightings Across Australia", x = "Longitude", y = "Latitude", subtitle = "1987 - Present") +
  theme_minimal()

combined_plot
```

**Temporal expectation**

We expect sightings of the Weeping Mulla Mulla to be more frequent from
March to October. This expectation aligns with our observations, as
there are fewer points on the map in January, February, November, and
December.

```{r}
combined_plot +
  facet_wrap(~monthName, ncol = 6) +
  theme_void() +
  labs(title = "Weeping Mulla Mulla Sightings By Month")
```

So far, the map data has shown us the areas where observations are most
frequent. We can see that there are primarily two clusters, one in
Western Australia and another in the Northern Territory. These clusters
are also evident on some of the maps by month.

**Trend expectation**

```{r}
weeping_mulla_mulla %>%
  group_by(Year) %>%
  count() %>%
  ggplot(aes(x = Year, y = n, fill = as.factor(Year))) +
  geom_histogram(stat = "identity") +
  scale_fill_manual(values = c("2020" = "#c599b2", "2021" = "#c599b2")) +
  labs(
    title = "Weeping Mulla Mulla Observations by Year",
    y = "Number of observations"
  ) +
  theme_minimal() +
  theme(legend.position = "none", legend.title = element_blank())
```

We can see that the yearly observations do not meet our expectation of
reduced observations during COVID-19. There was no decrease in
observations from 2000 to 2021.

## Summary

## References
